MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-23 Manuscript/Manuscript-2.Rmd
vsn 2

Evolutionary biologists commonly seek to describe the relative strength of phylogenetic signal in phenotypic traits, to determine the extent to which shared evolutionary history has influenced trait covariation among taxa. This is often accomplished by interpreting empirical estimates of $\lambda$; with $\lambda\approx0$ signifying 'weak' phylogenetic signal, and $\lambda\approx1$ being interpreted as 'strong' phylogenetic signal [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. Other approaches for interpreting $\lambda$ are more statistical, through the use of confidence intervals [@Vandelook2019] or likelihood ratio tests that compare the observed model fit to that obtained when $\lambda=0$ or $\lambda=1$ [@Freckleton_et_al2002; @Cooper2010; @Bose2019]. Likewise, qualitative comparisons of $\lambda$ across multiple phenotypic traits have also been used to infer whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019]. Indeed, it seems intuitive to interpret the strength of phylogenetic signal in this manner, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are well known ($\lambda=0$ is no phylogenetic signal, while $\lambda=1$ corresponds to Brownian motion). However, equating values of $\lambda$ directly to the strength of phylogenetic signal presumes two important statistical properties that have not been fully examined. \hfill\break

First, it presumes that values of $\lambda$ can be precisely estimated, and that parameter estimates of $\lambda$ are representative of the underlying phylogenetic signal. Understanding the precision in estimating levels of phylogenetic signal is paramount, because biological inferences regarding the strength of phylogenetic signal in phenotypic datasets requires high precision of its estimation. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise with greater sample sizes [@Cohen1988]. However, this conclusion also assumes that the precision in estimating $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while $\lambda$ is observed on a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\lambda$ must be a statistical effect size. Effect sizes are a measure the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets [e.g., meta-analysis: @HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared across datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunatley, not all statistical parameters and test statistics are effect sizes, and thus many summary measures must first be converted to standardized units (i.e., an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\lambda$ is a statistical effect size can comparisons of estimates across datasets be interpretable. To date, it has not been determined whether $\lambda$ is an effect size for phylogenetic signal.  \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We then derive a standardized effect size for phylogenetic signal, and apply the concept to two common measures of phylogenetic signal: $\lambda$ ($Z_{\lambda}$) and $Kappa$ ($Z_K$). In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. We also find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that this variation is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. Additionally, the same $\lambda_{est}$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. Likewise, the precision of $Z_{\lambda}$ is less consistent than that observed with $Z_K$, across the range of input values, implying that $Z_K$ is a more robust effect size measure. Further, because $Z_K$ is a standardized effect size, the strength of phylogenetic signal is more readily interpretable, and may be compared across datasets. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

>>>>

Evolutionary biologists commonly seek to describe the relative strength of phylogenetic signal in phenotypic traits, to determine the extent to which shared evolutionary history has influenced trait covariation among taxa. This is often accomplished by interpreting empirical estimates of $\lambda$; with $\lambda\approx0$ signifying 'weak' phylogenetic signal, and $\lambda\approx1$ being interpreted as 'strong' phylogenetic signal [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. Other approaches for interpreting $\lambda$ are more statistical, through the use of confidence intervals [@Vandelook2019] or likelihood ratio tests that compare the observed model fit to that obtained when $\lambda=0$ or $\lambda=1$ [@Freckleton_et_al2002; @Cooper2010; @Bose2019]. Likewise, qualitative comparisons of $\lambda$ across multiple phenotypic traits have also been used to infer whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019]. Indeed, it seems intuitive to interpret the strength of phylogenetic signal in this manner, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are well known ($\lambda=0$ is no phylogenetic signal, while $\lambda=1$ corresponds to Brownian motion). However, equating values of $\lambda$ directly to the strength of phylogenetic signal presumes two important statistical properties that have not been fully explored. \hfill\break

First, it presumes that values of $\lambda$ can be precisely estimated, and that parameter estimates of $\lambda$ are representative of the underlying phylogenetic signal. However, for such applications, understanding the precision in estimating levels of phylogenetic signal is paramount, because biological inferences regarding the strength of phylogenetic signal in phenotypic datasets requires high precision of its estimation. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise with greater sample sizes [@Cohen1988]. However, this conclusion also assumes that the precision in estimating $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while $\lambda$ is observed on a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\lambda$ must be a statistical effect size. Effect sizes are a measure the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets [e.g., meta-analysis: @HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared across datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunatley, not all statistical parameters and test statistics are effect sizes, and thus many summary measures must first be converted to standardized units (i.e., an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\lambda$ is a statistical effect size can comparisons of estimates across datasets be interpretable. For the case of $\lambda$, this has not yet been explored.  \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. We also find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that the precision in estimating $\lambda$ is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. Additionally, the same $\lambda_{est}$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. Thus, $\lambda$ is not a reliable effect size for measuring the strength of phylogenetic signal. We subsequently derive a standardized effect size for measuring the strength of phylogenetic signal in phenotypic datasets, and apply the concept to two common measures of phylogenetic signal: $\lambda$ and $Kappa$. Through simulations across a wide range of conditions, we find that the precision of effect sizes based on $\lambda$ ($Z_{\lambda}$) are less reliable than that those based on $Kappa$ ($Z_K$), implying that $Z_K$ is a more robust effect size measure. Additionally, we propose a two-sample test statistic that may be used to compare the strength of phylogenetic signal among datasets. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

### 2020-04-23 Manuscript/Manuscript-2.tex
vsn 2

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with \(\lambda\approx0\) signifying `weak' phylogenetic
signal, and \(\lambda\approx1\) being interpreted as `strong'
phylogenetic signal (e.g., De Meester et al. 2019; Pintanel et al. 2019;
Su et al. 2019). Other approaches for interpreting \(\lambda\) are more
statistical, through the use of confidence intervals (Vandelook et al.
2019) or likelihood ratio tests that compare the observed model fit to
that obtained when \(\lambda=0\) or \(\lambda=1\) (Freckleton et al.
2002; Cooper et al. 2010; Bose et al. 2019). Likewise, qualitative
comparisons of \(\lambda\) across multiple phenotypic traits have also
been used to infer whether the strength of phylogenetic signal is
greater in one trait as compared to another (e.g., Liu et al. 2019; Bai
et al. 2019). Indeed, it seems intuitive to interpret the strength of
phylogenetic signal in this manner, as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are well known (\(\lambda=0\) is no phylogenetic signal, while
\(\lambda=1\) corresponds to Brownian motion). However, equating values
of \(\lambda\) directly to the strength of phylogenetic signal presumes
two important statistical properties that have not been fully examined.
\hfill\break

First, it presumes that values of \(\lambda\) can be precisely
estimated, and that parameter estimates of \(\lambda\) are
representative of the underlying phylogenetic signal. Understanding the
precision in estimating levels of phylogenetic signal is paramount,
because biological inferences regarding the strength of phylogenetic
signal in phenotypic datasets requires high precision of its estimation.
One study (Boettiger et al. 2012) found that estimates of Pagel's
\(\lambda\) displayed less variation (i.e., greater precision) when data
were simulated on a large phylogeny (\(N=281\)) as compared to a small
one (\(N=13\)). From this observation it was concluded that insufficient
data (i.e., the number of species) was the underlying cause of the
increased variation across parameter estimates (Boettiger et al. 2012).
Indeed, such a pattern is common with statistical estimators, as summary
statistics and parameters are often more precise with greater sample
sizes (Cohen 1988). However, this conclusion also assumes that the
precision in estimating \(\lambda\) remains constant across its range
(\(\lambda = 0 \to 1\)); an assumption that to date, has not been
verified. Thus, despite widespread use of Pagel's (1999) \(\lambda\) in
macroevolutionary studies, at present, we still lack a general
understanding of the precision with which \(\lambda\) can estimate
levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while \(\lambda\) is observed on a bounded scale (\(0\to1\)),
this does not \emph{de-facto} imply that the estimated values of this
parameter correspond to the actual strength of the underlying input
signal in the data. For this to be the case, \(\lambda\) must be a
statistical effect size. Effect sizes are a measure the magnitude of a
statistical effect in data, represented on a common scale (Glass 1976;
Cohen 1988). Effect sizes have widespread use in many areas of the
quantiative sciences, as they represent measures that may be readily
summarized across datasets (e.g., meta-analysis: Hedges and Olkin 1985;
Glass 1976; Arnqvist and Wooster 1995), or compared across datasets
(e.g., Adams and Collyer 2016, 2019a). Unfortunatley, not all
statistical parameters and test statistics are effect sizes, and thus
many summary measures must first be converted to standardized units
(i.e., an effect size) for meaningful comparison (see Rosenthal 1994).
As a consequence, it follows that only if \(\lambda\) is a statistical
effect size can comparisons of estimates across datasets be
interpretable. To date, it has not been determined whether \(\lambda\)
is an effect size for phylogenetic signal. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We then
derive a standardized effect size for phylogenetic signal, and apply the
concept to two common measures of phylogenetic signal: \(\lambda\)
(\(Z_{\lambda}\)) and \(Kappa\) (\(Z_K\)). In general we find that while
PGLS parameters (e.g., \(\beta\)) are accurately estimated with the
inclusion of phylogenetic signal, estimates of \(\lambda\) are not. We
also find that estimates of \(\lambda\) vary widely for a given input
value of phylogenetic signal, and that this variation is not constant
across the range of input signal, with decreased precision when
phylogenetic signal is of intermediate strength. Additionally, the same
\(\lambda_{est}\) may be obtained from datasets containing vastly
different input levels of phylogenetic signal. Likewise, the precision
of \(Z_{\lambda}\) is less consistent than that observed with \(Z_K\),
across the range of input values, implying that \(Z_K\) is a more robust
effect size measure. Further, because \(Z_K\) is a standardized effect
size, the strength of phylogenetic signal is more readily interpretable,
and may be compared across datasets. We conclude that estimates of
phylogenetic signal using Pagel's \(\lambda\) are often inaccurate, and
thus interpreting strength of phylogenetic signal in phenotypic datasets
based on this measure is compromised. By contrast, effect sizes obtained
from \(Kappa\) hold promise for characterizing phylogenetic signal, and
for comparing the strength of phylogenetic signal across datasets.

>>>>

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with \(\lambda\approx0\) signifying `weak' phylogenetic
signal, and \(\lambda\approx1\) being interpreted as `strong'
phylogenetic signal (e.g., De Meester et al. 2019; Pintanel et al. 2019;
Su et al. 2019). Other approaches for interpreting \(\lambda\) are more
statistical, through the use of confidence intervals (Vandelook et al.
2019) or likelihood ratio tests that compare the observed model fit to
that obtained when \(\lambda=0\) or \(\lambda=1\) (Freckleton et al.
2002; Cooper et al. 2010; Bose et al. 2019). Likewise, qualitative
comparisons of \(\lambda\) across multiple phenotypic traits have also
been used to infer whether the strength of phylogenetic signal is
greater in one trait as compared to another (e.g., Liu et al. 2019; Bai
et al. 2019). Indeed, it seems intuitive to interpret the strength of
phylogenetic signal in this manner, as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are well known (\(\lambda=0\) is no phylogenetic signal, while
\(\lambda=1\) corresponds to Brownian motion). However, equating values
of \(\lambda\) directly to the strength of phylogenetic signal presumes
two important statistical properties that have not been fully explored.
\hfill\break

First, it presumes that values of \(\lambda\) can be precisely
estimated, and that parameter estimates of \(\lambda\) are
representative of the underlying phylogenetic signal. However, for such
applications, understanding the precision in estimating levels of
phylogenetic signal is paramount, because biological inferences
regarding the strength of phylogenetic signal in phenotypic datasets
requires high precision of its estimation. One study (Boettiger et al.
2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise with greater sample sizes (Cohen 1988). However,
this conclusion also assumes that the precision in estimating
\(\lambda\) remains constant across its range (\(\lambda = 0 \to 1\));
an assumption that to date, has not been verified. Thus, despite
widespread use of Pagel's (1999) \(\lambda\) in macroevolutionary
studies, at present, we still lack a general understanding of the
precision with which \(\lambda\) can estimate levels of phylogenetic
signal in phenotypic datasets. \hfill\break

Second, while \(\lambda\) is observed on a bounded scale (\(0\to1\)),
this does not \emph{de-facto} imply that the estimated values of this
parameter correspond to the actual strength of the underlying input
signal in the data. For this to be the case, \(\lambda\) must be a
statistical effect size. Effect sizes are a measure the magnitude of a
statistical effect in data, represented on a common scale (Glass 1976;
Cohen 1988). Effect sizes have widespread use in many areas of the
quantiative sciences, as they represent measures that may be readily
summarized across datasets (e.g., meta-analysis: Hedges and Olkin 1985;
Glass 1976; Arnqvist and Wooster 1995), or compared across datasets
(e.g., Adams and Collyer 2016, 2019a). Unfortunatley, not all
statistical parameters and test statistics are effect sizes, and thus
many summary measures must first be converted to standardized units
(i.e., an effect size) for meaningful comparison (see Rosenthal 1994).
As a consequence, it follows that only if \(\lambda\) is a statistical
effect size can comparisons of estimates across datasets be
interpretable. For the case of \(\lambda\), this has not yet been
explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that while PGLS parameters (e.g., \(\beta\)) are accurately estimated
with the inclusion of phylogenetic signal, estimates of \(\lambda\) are
not. We also find that estimates of \(\lambda\) vary widely for a given
input value of phylogenetic signal, and that the precision in estimating
\(\lambda\) is not constant across the range of input signal, with
decreased precision when phylogenetic signal is of intermediate
strength. Additionally, the same \(\lambda_{est}\) may be obtained from
datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable effect size for measuring
the strength of phylogenetic signal. We subsequently derive a
standardized effect size for measuring the strength of phylogenetic
signal in phenotypic datasets, and apply the concept to two common
measures of phylogenetic signal: \(\lambda\) and \(Kappa\). Through
simulations across a wide range of conditions, we find that the
precision of effect sizes based on \(\lambda\) (\(Z_{\lambda}\)) are
less reliable than that those based on \(Kappa\) (\(Z_K\)), implying
that \(Z_K\) is a more robust effect size measure. Additionally, we
propose a two-sample test statistic that may be used to compare the
strength of phylogenetic signal among datasets. We conclude that
estimates of phylogenetic signal using Pagel's \(\lambda\) are often
inaccurate, and thus interpreting strength of phylogenetic signal in
phenotypic datasets based on this measure is compromised. By contrast,
effect sizes obtained from \(Kappa\) hold promise for characterizing
phylogenetic signal, and for comparing the strength of phylogenetic
signal across datasets.
